---
title: "Collect radial measurements from a calcified structure by interactively selecting annuli"
author: "Derek H. Ogle"
date: "`r Sys.Date()`"
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: vignette
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Collect radial measurements from a calcified structure by interactively selecting annuli}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r eval=FALSE, echo=FALSE}
## Run this code to actually build the vignette for the package
devtools::build_vignettes()
devtools::install()
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
The size of fish at a previous time in their life is often estimated through a method called "back-calculation".[^VigliolaMeekan] Back-calculation of previous lengths requires accurate measurements of annual growth on calcified structures (hereafter, structures) extracted from individual fish and a suitable model that relates growth on the structure to growth of the fish. The [FishBC software](https://fisheries.org/bookstore/all-titles/software/70317/) is commonly used to measure lengths on the structure and apply a back-calculation model to estimate size at a previous age. This `RFishBC` package was developed to replicate the functionality of the FishBC software because FishBC only works on out-dated computers, there are no plans to update it, and it is not open source. Methods for making measurements on structure images is demonstrated in this vignette. Using those measurements to back-calculate fish length at a previous age is demonstrated in [this separate vignette](backcalculation.html).

This vignette assumes that you have a basic understand of back-calculation and the data (especially radial measurements made to annuli on a calcified structure) required to make back-calculations.

The methods described in this vignette require the following packages.

```{r message=FALSE}
library(RFishBC)
library(dplyr)    # for mutate(), inner_join()
```

\  

\  

----

# Collect Radii Data
The `digitizeRadii()` function is used to

1. Load an image (JPG, PNG, or BMP) of a structure.
1. Optionally provide a scaling for the image.
1. Interactively select annuli on the structure image.
1. Create an R data object file (i.e., ".Rdata" file) that contains, among other things, the radial measurements between the selected annuli.

For example (*but described thoroughly further below*), the line below loads the structure image in "Scale_1.png" (found in the current working directory), allows the user to select points that represent annuli (noting that the edge is not considered to be an annulus; this is default behavior), identifies that this reading of the image should be labelled as "DHO", and saves the results in an R data object file called "Scale_1_DHO.Rdata". Note that the radial measurements will be recorded as proportions of the image scale (i.e., 0 to 1) because no scaling factor was given or scale bar was processed.

```{r eval=FALSE}
fn <- digitizeRadii("Scale_1.png",ID="1",reading="DHO")
```

As this process requires interactive input from the user it is not shown in this static vignette. It should be noted, however, that reminders for the user are printed in the console at each step of the process.

Given the number of tasks required by `digitizeRadii()` it has many arguments. These are discussed in detail below.

### The Basics
The first five arguments to `digitizeRadii()` are related to information about the structure. The first argument is the image file name, which must be fully pathed if the image file is not in the current working directory. If this argument is not given, then a dialog box will be provided from which the file can be selected. A unique identifier for the fish/structure **must** be given in `ID=` and a unique label for the readings **must** be given in `reading=`.[^readings]

The R data object that will ultimately be created by `digitizeRadii()` will have the same name as the image file[^excludeextension] but including the suffix optionally provided in `suffix=`. If nothing is given in `suffix=` then the value in `reading=` will be used instead. For example, if the structure image file was named "Scale_1.png" then the resultant R data object file will be named "Scale_1_DHO.Rdata" if `reading="DHO"` or "Scale_1_TESTING.Rdata" if `reading="DHO"` and `suffix="TESTING"`.

The user may *optionally* provide more description for the structure in `description=`, which will be saved in the R data file object and shown in various plots and tables. For example, one may use `description="Cisco scale read once by Ogle on 22-Apr-18"` to provide more information about the structure reading.

By default, `digitizeRadii()` opens each image file in a separate window. The use of the separate window helps avoid issues with the RStudio Plots pane and is especially useful when working on a system with two monitors.[^twomonitors] This window is also configured so that when you resize the window, the shape of the structure image (i.e., its aspect ratio) will not change. One can suppress showing the structure image in a separate window with `sepWindow=FALSE` in `digitizeRadii()`. For example, using this option for subsequent images will reduce the number of windows that are open at any one time.

### Setting the Scale on the Image
**Scale-bar On Image:** Several back-calculation methods (see [this vignette](backcalculation.html)) require estimating the relationship between fish and structure length. For those methods, the actual lengths of the structure and each radial measurement are required. Collection of actual lengths, rather than proportional lengths, requires a scaling factor to convert measurements on the image to actual lengths of the structure. If a "scale bar" of known length exists on the image, then, in `digitizeRadii()`, use `scaleBar=TRUE` and give the actual length of the scale bar to `scaleBarLength=`. You will then be prompted to select the two end points of the scale bar on the structure image prior to selecting points that represent annuli. An appropriate scaling factor will then be computed from your selections and the radial measurements will be converted to the actual scale (see further below).

**Separately Define Scaling Factor:** In some applications it may not be possible to obtain an image with a scale bar. In these instances the user may capture the image at a selected magnification of the microscope and then capture an image of an object of known length at that same magnification. A scaling factor may then be computed from the image with the object of known length and applied to the structure image. Once this scaling factor is known, it can be given to `scalingFactor=` in `digitizeRadii()`.

The `findScalingFactor()` function can be used to load the image that contains an object of known length on which the two end points of the image can be selected. The value returned from this function is the scaling factor that can then be given to `scalingFactor=` in `digitizeRadii()`.[^sforder] The `findScalingFactor()` function requires the name of the image with the object of known length as the first argument (if no name is given then you will be asked to browse to the image) and the known length of the object in `knownLength=`.

If no `scalingFactor=` is given or derived from a scale bar, then the radial measurements returned by `digitizeRadii()` are simply proportional to the unknown actual lengths on the structure.

### Selecting Annuli on the Image
Finally, the user will be asked to select points on the structure image that represent annuli. However, the point that represents the focus must be selected first, followed by the structure margin/edge.[^invisiblepoints] If `addTransect=TRUE` (the default behavior) then a a linear transect will be drawn between these two points to add marking the other points. The color, width, and type of the transect can be modified with `col.transect=`, `lwd.transect=`, and `lty.transect=`, respectively.

When the last point has been selected, the user must press the ESCape key in Windows or with Mac OS X (or, in Windows, select STOP in the menu on the structure image window). The plotting character, color, and relative size of the selected points can be changed from the default settings with `pch.pts=`, `col.pts=`, and `cex.pts=`.

If the distance between the last selected annular point and the structure margin does NOT represent a full season's of growth then use `edgeIsAnnulus=TRUE` (**not** default behavior) in `digitizeRadii()` to highlight this measurements as incomplete growth and assure proper processing of the selected points.

\  

\  

----

# Examine Radii Data
### Data.frame of Radii
The result from the previous section is an R object data file in the current working directory. Among other things, this data file contains a data.frame of radial measurements to successive annuli. This data.frame can be extracted from the data file with `combineData()` using the name of the R object data file as the first argument.
```{r eval=FALSE}
df <- combineData("Scale_1_DHO.RData")
```
```{r echo=FALSE}
## This loads the file from the external data folder
fn <- "Scale_1_DHO.RData"
load(fn)
df <- combineData(fn)
```
```{r}
df
```

The `combineData()` function is more interesting when considering multiple structures (see next section).

### View Annular Markings
One can review the markings on the structure with `showDigitizedImage()`, which requires only the name of an R data object file name.[^sdifilename] The plotting character, color, and relative size of the selected points may be changed with `pch.pts=`, `col.pts=`, and `cex.pts=`. The color and width of the "transect" may be changed with `col.transect=` and `lwd.transect=` or excluded altogether with `showTransect=FALSE`.

```{r}
showDigitizedImage("Scale_1_DHO.RData")
```

\  

\  

----

# Examine Multiple Reads of the Same Structure
In some instances, one may be interested in visually comparing the selected points from multiple reads of the same structure. For example, the code below collects a second set of points for the "Scale_1.png" image.[^noteabove] Note the different value in `reading=`.
```{r eval=FALSE}
digitizeRadii("Scale_1.png",ID="1",reading="DHO2")
```

As shown above, `showDigitizedImage()` can be used to overlay the selected points onto the structure image. A second set of points can be overlaid onto the active plot by including `add=TRUE` in a second call to `showDigitizedImage()` with a second R object data file name constructed from the same image. For example, the first line below overlays the annuli from the first selections onto the image. The second line then overlays the annuli from the second selections onto this (because `add=TRUE`) and uses a different color for the transect and points.
```{r}
showDigitizedImage("Scale_1_DHO.RData")
showDigitizedImage("Scale_1_DHO2.RData",add=TRUE,
           col.transect="red",col.pts="yellow")
```

\  

\  

----

# Combine Data from Multiple Structures
### File Organization
Of course, most analyses will consist of collecting radial measurements from structures from many fish. In this section, I demonstrate how to combine measurements from multiple structures. This demonstration assumes that all structure image files of interest are in the working directory.[^workingdirectory]

The `listFiles()` function may be used to identify all filenames in the current working directory that have a the file extension given in the first argument. It may be more efficient to save these filenames into an object (e.g., `fns`) as shown below.
```{r}
( fns <- listFiles("png") )
```

### Processing Multiple Structure Images
The user would then use the techniques described for one structure image in the first section above for **each** of the structure images in the working directory. This will require multiple calls to `digitizeRadii()`, each with a different structure image file name. If the file names are stored in an object as described above, then these names can be extracted by their position in the object. For example, the two lines below would be used to construct radii data from the images in the first two positions of the `fns` object (e.g., `fns[1]` and `fns[2]`). Note that the `ID=` argument should change to represent the unique identification number of the fish.
```{r eval=FALSE}
digitizeRadii(fns[1],ID="1",reading="DHO")
digitizeRadii(fns[2],ID="2",reading="DHO")
```

### Combining Data Extracted from Multiple Structure Images
Once all structure image files have been processed, the radial measurements can be combined into one data.frame with `combineData()`. The first argument to this function is a vector of R object file names, which can be constructed using `listFiles()` with the `.RData` extension used for R object file names).
```{r results='hide'}
( fns2 <- listFiles("RData") )
```
```{r echo=FALSE}
fns2 <- fns2[-2]
fns2
```
```{r}
dfrad <- combineData(fns2)
dfrad
```

### Preparing for Back-Calculations
Other information about the fish (e.g., location of capture, length, sex) is likely held in a separate file. Below, example "other" data are loaded into the `dffish` data.frame. Note that the `ID` and `reading` variables created from processing the structure images above are characters. In this case, `read.csv()` reads the `ID` variable from the external data file as numeric (because the unique IDs were simple numbers). The second line of code below converts these numeric IDs to characters so that this data.frame can be joined with the radial measurements data.frame from above.[^wrongtypesinjoin]
```{r}
dffish <- read.csv("FishData.csv") %>%
  mutate(ID=as.character(ID)) %>% 
  inner_join(dfrad,by="ID")
dffish
```

\  

\  

----

# Footnotes

[^VigliolaMeekan]: See [Vigliola and Meekan (2009)](https://www.researchgate.net/publication/226394736_The_Back-Calculation_of_Fish_Growth_From_Otoliths) for background information.

[^readings]: The `readings` item is primarily used when the structure is read more than once. However, it is a required component in this package, whether multiple reads of the same structure are planned or not.

[^excludeextension]: The extension (e.g., ".png" or "jpg") will not be included in the resultant R data object file.

[^twomonitors]: For example, you could have R/RStudio on one monitor and the window with the structure image on the other.

[^sforder]: Thus, `findScalingFactor()` would be run prior to `digitizeRadii()`.

[^invisiblepoints]: When the structure image is in the "Plots" pane of RStudio, the points will not be visible until after you have finished selecting all points.

[^sdifilename]: If no file name is given, then a dialog box will appear from which the data file can be selected.

[^noteabove]: The first set of readings were made in the first section above.

[^workingdirectory]: This process will be more efficient if not other image files of the same sort (e.g., ".png") are in this directory. The working directory may be seen with `getwd()` and set with `setwd()` or any of the items under the "Session" menu and "Set Working Directory" submenu items in RStudio.

[^wrongtypesinjoin]: The two files cannot be joined by the "ID" variable if the "ID" variable is of different types (e.g., character and numeric) in the two files. Thus, as shown here, they must be coerced to be the same type. This line of code would no be needed if the "ID" variables were of the same type in the two files as would likely occur if the "ID" variable was not simply numbers.